Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2020-05-18 submit_v03/manuscript.tex
m

The power law and exponential statistical distributions are among the most common patterns found in natural phenomena. For example, the distribution of the frequency of words in many languages is known to follow a Zipf’s law (which is a  power law). These distributions arise any time a phenomenon is characterised by commonly occurring small events and rarely produced large events \citep[e.g. ][]{adamic_zipf_2000}. Zipf law also describes population ranks of cities in various countries, corporation sizes, income rankings, ranks of the number of people watching the same TV channel, etc. The magnitudes of earthquakes, hurricanes, volcanic eruptions and floods; the sizes of meteorites or the losses caused by business interruptions from accidents, are also well described by power laws \citep{sornette_critical_2006,  benguigui_classificacion_2016}. The power law behaviour has been observed in a variety of systems.  These include, for example, stock market fluctuations, sizes of computer files or word frequency in languages \citep{mitzenmacher_brief_2004, newman_power_2005, simkin_theory_2006}. Power laws have also been widely used in biological sciences.  Some examples are the analysis of connectivity patterns in metabolic networks \citep{jeong_large_2000} or the number of species observed per unit area in ecology \citep{martin_origin_2006, frank_common_2009}. More examples can be found in the literature \citep{martin_origin_2006, maccone_KLT_2010, barabasi_scale_2009, maccone_evolution_2014,  maccone_lognormals_2014, benguigui_classificacion_2016}.

@@

In this Section we discuss the degree of complexity in the model,
considering a compromise between the accuracy of the model and the
number of parameters that are free or with a high uncertainty.
%
Firstly, we emphasize that the odds of a causal contact between two
\cetis{} should not be considered as the odds of a contact between two
intelligent civilizations, and in fact the later could be much lesser
than the former.
%
Indeed, in order to establish a contact between any two entities, a
minimum degree of compatibility must be accomplished without any
previous agreement, making the possibility of a contact with a message
that could be deciphered highly rare \citep[see e.g. ][]{forgan_collimated_2014}.
%
Besides the trade--off between the simplicity and the complexity of
the experiments, further analysis could be performed following this
framework in order to explore possible implications of the results for
more detailed configurations.
%
For example, the communication method (isotropic, collimated,
serendipitous) can affect the observables, making it necessary to
implement a correction factor.
%
Taking this into account, our results regarding the probability of
causal contact should be considered as upper limits for effective
contacts, since they depend on the efficiency of both the emitter and
the receiver to broadcast and scan the sky for intelligent signals,
respectively \citep{grimaldi_signal_2017}.
%
A correction by coverage ratio in the detection and by a targeting
ratio in the emission could be easily implemented in the simulation,
although the effect of reducing the probability of contact is
basically the product of the efficiency ratios and thus such
implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a
constant correction factor equal to the combined emission/reception
efficiency \citep{smith_broadcasting_2009, anchordoqui_upper_2019,
forgan_collimated_2014} or for beam--like transmissions
\citep{grimaldi_signal_2017}.
%
Other considerations include the effects of alignments, the use of
stars as sources or amplifiers \citep{Edmondson2003, borra_searching_2012}, 
the nature of the message carrier, the
use of a spatial distribution that actually resembles the spiral shape
and the width of the disc of the Galaxy, different distribution
functions for the mean lifetime of \cetis{}, and different
efficiencies for the sphere of the causal region of each \ceti{}
resembling different searching strategies
\citep{hippke_interstellar_2017}.
%
It is also possible to consider that D$_{max}$ is different for
different \cetis{}.  For example a power law where powerful emissions
are rare and weak emissions are common could be an improvement to the
model.
%
In this work we chose not to implement this for the sake of
simplicity.
%
Finally, the role of the message contents could influence on the
lifespan of a \ceti{} that receives a message, although the
implementation of such behavior would increase the number of free
parameters and would be highly speculative.

@@

Given a parametric model $M(\tau_a, \tau_s, D_{max})$, which includes
the functional forms of the statistical distributions and fixed values
for the hypotheses, a discrete event simulation is performed by
keeping track of a set of variables that change each time a meaningful
event is produced.
%
The main variables that follow the evolution of the simulation are the
positions of stars, which are sampled randomly within the GHZ, the
time of the awakening of each \ceti{} (A event), and the time of the
vanishing of each \ceti{} (D event).
%
The variables that can be deduced from the previous ones include the
number of \cetis{} in casual contact with at least another \ceti{} at a given
time, the number of \cetis{} as a function of time, the number of
\cetis{}
that receive a message at least one time, the number of \cetis{} that
receive a message at least one time and successfully deliver an answer
and the number distribution of waiting time to receive a message.
%
All these quantities are updated each time one of the four events (A,
B, C, D) is produced.

@@

Since the method is based on events, the first step is to define an
architecture of events, the relationships between them and how events
trigger changes on the state of the system,
so that the system is updated each time an event is produced.
%
We consider a simulated system that represents the galactic habitable
zone which, on a first approach, is a 2-dimensional annular region.
%
The adopted values for the geometry of the GHZ are 20 kly and 60 kly, for the inner
and outer radius, respectively \citep{lineweaver_galactic_2004}.
% ojo, ese paper dice 7-9 kpc, o sea tiene un limite superior mucho
% mas chico...
%
This simple model does not take into account the variations in stellar
density given by the spiral structure.
%
Although there are several possible approaches, we chose to follow the
evolution of the system according to the following events:

@@

We have chosen this nomenclature so that it can be easily remembered
from the initial A, B, C, D.
%
The \blackout is produced when one of the two \cetis{} halts in its
capability of emitting and receiving signals.
%
It can occur at the same time the \doomsday or before.
%
Similarly, a \ccontact can be produced at the time of the \aawakening
or later, and there can be several \contacts for the same node.
%
The time marks for each event (A and D for each \ceti{} and C and B
for each causal contact) are stored as a result of the simulation.
%
Also, the list of active \cetis{} is obtained as a function of time.
%
%Some fixed variables must be set in order to carry out the simulation:
%the size of the Galactic Habitable Zone, the mean lifetime of a
%\ceti{}, $\tau_s$, the mean waiting time for the appearance of another
%\ceti{}, $\tau_a$, and the maximum distance a message can be detected.

The Galactic Habitable Zone is set at fixed values of the inner and outer
radii, radial symmetry is assumed, and the with of the galactic disk
is neglected compared to the radial size.
%
The two radii are measured in light years, with the aim to maintain a
single and comprehensive unit for both space and time coordinates. 
%
%}}}

@@

%Quality assurance
%
Regarding the quality assurance of the software, we endorse the PEP-8
style guide for python (van Rossum et al. 2001) throughout the entire
project, which improves the code maintainability and helps to minimise
the so called ''code-entropy'' of future versions.
%
The deviations in style were analysed with the flake8 tool, which
improves the readability and consistency of the code.
%
Testing and code coverage have been applied to the package and the
development process as standard qualitative and quantitative metrics.
%
The different parts and functions can be validated to assert if they
return the expected results on specific test cases and testing
scenarios \citep{jazayeri_engeneering_2007}.
%
On the other hand, code coverage measures the fraction visited and
executed in the code by the testing procedure
\citep{miller_testing_1963}.
%
The advantage of an exhaustive code-coverage is that it ensures that
most of the sources of errors have been considered.
%
In the \hs{} project we provide test tools for the distinct parts of
the simulation setup and run, reaching a 99\% of code coverage in the
version used for this work.
%
For every new version of the code, all these quality assurance methods
and metrics are run automatically by a continuous-integration service.
%
It is possible to reproduce the experiments performed in this work
using the \hs{} package.
%
The installation is available through a git repository and also on the
Python Package Index (PyPI).
%
Since the simulation setup is configurable, the time required for a
simulation to complete depends on the simulation parameters.

>>>>

The power law and exponential statistical distributions are among the most common patterns found in natural phenomena.
%
For example, the distribution of the frequency of words in many languages is known to follow a Zipf’s law (which is a 
power law).
%
These distributions arise any time a phenomenon is characterised by commonly occurring small events and rarely produced
large events \citep[e.g. ][]{adamic_zipf_2000}.
%
Zipf law also describes population ranks of cities in various countries, corporation sizes, income rankings, ranks of
the number of people watching the same TV channel, etc.
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and floods; the sizes of meteorites or the losses caused
by business interruptions from accidents, are also well described by power laws
\citep{sornette_critical_2006}.
%
%% , benguigui_classificacion_2016}.
The power law behaviour has been observed in a variety of systems.
%
These include, for example, stock market fluctuations, sizes of computer files or word frequency in languages
\citep{mitzenmacher_brief_2004, newman_power_2005, simkin_theory_2006}. 
%
Power laws have also been widely used in
biological sciences.
%
Some examples are the analysis of connectivity patterns in metabolic networks \citep{jeong_large_2000} or the number of
species observed per unit area in ecology \citep{martin_origin_2006, frank_common_2009}.
%
More examples can be found in the literature 
\citep{martin_origin_2006, maccone_KLT_2010}
\citep{barabasi_scale_2009, maccone_evolution_2014}
\citep{maccone_lognormals_2014}

@@

In this Section we discuss the degree of complexity in the model, considering a compromise between the accuracy of the
model and the number of parameters that are free or with a high uncertainty.
%
Firstly, we emphasise that the odds of a causal contact between two \cetis{} should not be considered as the odds of a
contact between two intelligent civilisations, and in fact the latter could be much lesser than the former.
%
Indeed, in order to establish a contact between any two entities, a minimum degree of compatibility must be accomplished
without any previous agreement, making the possibility of a contact with a message that could be deciphered highly rare
\citep[see e.g. ][]{forgan_collimated_2014}.
%
Besides the trade--off between the simplicity and the complexity of the experiments, further analysis could be performed
following this framework in order to explore possible implications of the results for more detailed configurations.
%
For example, the communication method (isotropic, collimated, serendipitous) can affect the observables, making it
necessary to implement a correction factor.
%
Taking this into account, our results regarding the probability of causal contact should be considered as upper limits
for effective contacts, since they depend on the efficiency of both the emitter and the receiver to broadcast and scan
the sky for intelligent signals, respectively \citep{grimaldi_signal_2017}.
%
A correction by coverage ratio in the detection and by a targeting ratio in the emission could be easily implemented in
the simulation, although the effect of reducing the probability of contact is basically the product of the efficiency
ratios and thus such implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a constant correction factor equal to the combined
emission/reception efficiency \citep{smith_broadcasting_2009, anchordoqui_upper_2019, forgan_collimated_2014} or for
beam-like transmissions \citep{grimaldi_signal_2017}. 
%
Other considerations include the effects of alignments, the use of stars as sources or amplifiers \citep{Edmondson2003,
borra_searching_2012}, the nature of the message carrier.
%
Another improvement could be the use of a spatial distribution that resembles the spiral shape and the width of the disc
of the Galaxy.
%
Regarding the distributions of the parameters, different distribution functions for the mean lifetime of \cetis{} can be
implemented as alternatives.
%
Finally, regarding the degree of idealisation, the model admits different efficiencies for the sphere of the causal
region of each \ceti{}, featuring different searching strategies \citep{hippke_interstellar_2017}. 
%
It is also possible to consider that $D_{max}$ is different for different \cetis{}.
%
For example, a power law where powerful emissions are rare and low emissions are common could be an improvement to the
model.
%
In this work we chose not to implement this for the sake of simplicity.
%
Finally, the role of the message contents could influence on the lifespan of a \ceti{} that receives a message, although
the implementation of such behaviour would increase the number of free parameters and would be more speculative.

@@

A discrete event simulation is performed for a given model $M(\tau_a, \tau_s, D_{max})$ by keeping track of a set of variables that change each time an event happens. 
%
In addition, the model comprises elements that are fixed for all simulations, for example, the functional forms of the statistical distributions and the adopted values of particular variables (see Table XXX).
%
The main variables that follow the evolution of the simulation are: the positions of stars, which are sampled randomly within the GHZ; the time of the awakening of each \ceti{} (A event); and the time of the vanishing of each \ceti{} (D event). 
%
The variables that can be deduced from the previous ones include: the number of \cetis{} in casual contact with at least another \ceti{} at a given time; the number of \cetis{} as a function of time; the number of \cetis{} that receive a message at least one time; the number of \cetis{} that receive a message at least one time and successfully deliver an answer; and the number distribution of waiting time to receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D) occurs.

@@

Since the method is based on events, the first step is to define the relationships between them and how events trigger changes on the state of the system.
%
Once an event is produced, the system is updated by saving the corresponding change in the related variables.
%
For example, a B-type event diminishes the number of active \cetis{} by one, and also could affect the number of ongoing communications.
%
We consider a first approach to the galactic habitable zone as a 2-dimensional annular region.
%
The adopted values for the geometry of the GHZ are 2 kpc and 8 kpc, for the inner and outer radius, respectively \citep{lineweaver_galactic_2004}. 
% ojo, ese paper dice 7-9 kpc, o sea tiene un limite superior mucho
% mas chico...
%
This simple model does not take into account the variations in stellar density given by the spiral structure of the Galaxy.
%
The main task in setting a DE simulation is to define the events and what changes do these events produce on the system.
%
Although there are several possible approaches, we chose to follow the evolution of the system according to the following events:

@@

We have chosen this nomenclature so that it can be easily remembered
from the initials A, B, C, D.
%
The \blackout is produced when one of the two \cetis{} halts in its
capability of emitting and receiving signals.
%
It can occur at the same time the \doomsday or before.
%
Similarly, a \ccontact can be produced at the time of the \aawakening
or later, and there can be several \contacts for the same node.
%
The time marks for each event (A and D for each \ceti{} and C and B
for each causal contact) are stored as a result of the simulation.
%
Also, the list of active \cetis{} is obtained as a function of time.
%
%Some fixed variables must be set in order to carry out the simulation:
%the size of the Galactic Habitable Zone, the mean lifetime of a
%\ceti{}, $\tau_s$, the mean waiting time for the appearance of another
%\ceti{}, $\tau_a$, and the maximum distance a message can be detected.

% The Galactic Habitable Zone is set at fixed values of the inner and outer
% radii, radial symmetry is assumed, and the with of the galactic disk
% is neglected compared to the radial size.
% %
% The two radii are measured in light years, with the aim to maintain a
% single and comprehensive unit for both space and time coordinates. 
%
%}}}

@@

%Quality assurance
%
Regarding the quality assurance of the software, we endorse the PEP-8
style guide for python (van Rossum et al. 2001) throughout the entire
project, which improves the code maintainability and helps to minimise
the so called ''code-entropy'' of future versions.
%
The deviations in style were analysed with the flake8 tool, which
improves the readability and consistency of the code.
%
Testing and code coverage have been applied to the package and the
development process as standard qualitative and quantitative metrics.
%
The different parts and functions can be validated to assert if they
return the expected results on specific test cases and testing
scenarios \citep{jazayeri_engeneering_2007}.
%
On the other hand, code coverage measures the fraction visited and
executed in the code by the testing procedure
\citep{miller_testing_1963}.
%
The advantage of an exhaustive code-coverage is that it ensures that
most of the sources of errors have been considered.
%
In the \hs{} project we provide test tools for the distinct parts of
the simulation setup and run, reaching a 99\% of code coverage in the
version used for this work.
%
For every new version of the code, all these quality assurance methods
and metrics are run automatically by a continuous-integration service.
%
It is possible to reproduce the experiments performed in this work
using the \hs{} package.
%
The installation is available through a git repository and also on the
Python Package Index (PyPI).
%
Since the simulation setup is configurable, the time required for a
simulation to complete depends on the simulation parameters.
%
It also depends on the hardware that is used to make the run, and on whether the parallel option is set.
%
However, it is simple to make a local run of limited versions of the experiments to have a sense of the time required by
the code to complete the simulations.
%
More information on this can be found on the documentation of the software, available at
\url{https://hearsay.readthedocs.io/en/latest/}.
